---
title: "3 · Session title"
subtitle: "The one-line promise of this session"
---

<!--
Session page template. Ten parts, in order. See references/teaching-pages.md.

Parts 1 and 2 are generated from course.yml. Never type a time into this file.
Parts 3 to 10 are authored. Delete any part that does not apply, but keep the
order of the ones that remain.

For stack: none, delete the code chunks and keep everything else.
-->

::: {.session-meta}
::: {.meta-item}
[Teaching]{.meta-label}[11:15 – 12:00]{.meta-value}
:::
::: {.meta-item}
[Hands-on]{.meta-label}[12:00 – 13:00]{.meta-value}
:::
::: {.meta-item}
[Faculty]{.meta-label}[Name]{.meta-value}
:::
::: {.meta-item}
[Slides]{.meta-label}[[Open deck ↗](../slides/day1/s3.html){target="_blank"}]{.meta-value}
:::
:::

```{r}
#| include: false
# Hidden setup. Exists so the visible chunks below show only the line being
# taught. Loads the stack, sets a plot theme, reads the data.
library(tidyverse)
theme_set(theme_minimal(base_size = 12))
students <- read_csv("data/students.csv", show_col_types = FALSE)
```

## Learning objectives

About five, each starting with a verb the participant can do at the end.
Not "understand".

1. Read a dataset in and inspect it before touching it.
2. Name the three things every plot needs.
3. Build a histogram, a boxplot and a scatterplot.
4. Split a plot by a grouping variable.
5. Label a figure so that it stands alone in a manuscript.

## Start by looking

Worked example first, prose after. A beginner reads the output and then wants
to know why; explaining before showing gives them nothing to attach it to.

```{r}
glimpse(students)
```

240 rows, one per student. Note that `posttest` has missing values. We will
meet them shortly.

## The idea in one line

> data, plus a mapping from variables to visual channels, plus a geometry
> that draws it

```{r}
#| eval: false
ggplot(students, aes(x = pretest, y = posttest)) +
  geom_point()
```

Then two or three lines saying what each part did.

## A trap worth stopping on

```{r}
#| warning: true
ggplot(students, aes(x = posttest)) +
  geom_histogram(binwidth = 5)
```

::: {.callout-important}
## That warning is the point
`Removed 11 rows containing non-finite values`. Eleven students have no
post-test score. The message is easy to scroll past. Real analyses die here.

Read the warning, then decide explicitly what to do about the missing eleven,
and write that decision into your script.
:::

The count is fixed by the data generator, so it is safe to quote in prose.
See references/course-materials.md.

# Hands-on {#hands-on}

Level one, with an explicit anchor, so the schedule page can deep-link
`s3.qmd#hands-on` and a participant arriving at 12:00 lands here rather than at
the top of the page.

It is 12:00. Work in pairs. Raise a hand rather than stay stuck.

::: {.callout-note title="Exercise 1: Distribution of post-test scores"}
**Time:** 5 minutes

Draw a histogram of `posttest`. Choose a binwidth that shows the shape without
inventing detail.
:::

::: {.callout-tip title="Solution 1" collapse="true"}
```{r}
#| eval: false
ggplot(students, aes(x = posttest)) +
  geom_histogram(binwidth = 5)
```

The distribution is close to symmetric, so a mean is a fair summary here. The
eleven removed rows are the same eleven students as before.

A solution carries the interpretation, not only the code. Code answers "what do
I type". The participant came for "what does it mean".
:::

::: {.callout-note title="Exercise 2: Does the arm matter?"}
**Time:** 8 minutes

Plot `posttest` by `teaching_arm`. Show the raw points as well as the summary.
:::

::: {.callout-tip title="Solution 2" collapse="true"}
```{r}
#| eval: false
ggplot(students, aes(x = teaching_arm, y = posttest)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3)
```

The two arms overlap almost completely. There is no visible difference here,
and that is the honest answer.

Write at least one exercise per course whose answer is that the effect is
absent. Every exercise finding a clean result teaches that analysis always
finds something, which is the opposite of the lesson.
:::

## Key takeaways

::: {.callout-note appearance="minimal"}
- Every plot is data, a mapping, and a geometry.
- A warning about removed rows is a finding, not noise.
- Show the points when the sample is small enough to allow it.
:::

Each bullet is a sentence the participant could say out loud, not a topic name.

## Next

→ [4 · Next session](s4-next.qmd) : write a document that contains its own code.
